---
title: "Data Visualization: Part 1"
author: "YC Huang"
output: 
    github_document:
        toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
The data is obtained from [OSMI's Survey on Mental Health in the Tech Workplace in 2014](https://www.kaggle.com/osmi/mental-health-in-tech-survey) and is available as a csv file.


## Setting up the Workspace

We will be using the following packages. 
```{r libs and options, message=FALSE, warning=FALSE}
library(dplyr)     # data manipulation
library(forcats)   # handle factor variables
library(ggplot2)   # plotting
library(gridExtra) # arrange multiple ggplots
library(purrr)     # map functions
```

## Loading the Data

We had previously cleaned the data and saved it as a RDS file, so we will load the data directly:

```{r load data}
survey_clean <- readRDS("data/survey.rds")
```

## Visualizing Demographic Variables (Age, Gender, Country, State)

```{r echo=FALSE}
theme_set(theme_bw())
```

Based on the `Age` variable, the majority of respondents are within 25-35 years of age. The distribution of respondents is approximately normal, with a little skew towards ages 50 and above. Across `Gender`, about 80% of respondents taking this survey identified as male, alongside females and other genders:

```{r demo1}
plot_age <- ggplot(survey_clean, aes(x = Age)) +
    geom_bar() +
    scale_x_continuous(breaks = seq(20, 70, 10))
plot_gender <- ggplot(survey_clean, aes(x = Gender)) +
    geom_bar()
grid.arrange(plot_age, plot_gender, ncol = 1)
```

To better visualize the distributions for `Country` and `state`, we group the less frequent responses together using `fct_lump()`, before plotting the data:

```{r}
survey_clean %>% 
    # group less frequent responses together
    mutate(Country = fct_lump(Country, n = 9)) %>% 
    group_by(Country) %>% 
    summarise(total = n()) %>% 
    # reordering the country variable so that the countries appear in order of
    # least to most frequent (upward)
    mutate(Country = fct_reorder(Country, total)) %>%
    ggplot(aes(x = Country, y = total)) +
        geom_bar(stat = "identity") +
        coord_flip()
```

```{r}
survey_clean %>% 
    # remove NA responses
    filter(!is.na(state)) %>% 
    mutate(us_state = fct_lump(state, n = 15)) %>% 
    group_by(us_state) %>%
    summarise(total = n()) %>%
    mutate(us_state = fct_reorder(us_state, total)) %>%
    ggplot(aes(x = us_state, y = total)) +
        geom_bar(stat = "identity") +
        coord_flip()
```

From the plots above, we can see that a large majority of respondents (`r sum(survey_clean$Country == "United States")`) are from the United States. Among respondents who indicated their US state, most of them came from California.

## Visualizing Survey Responses (Single Variable)

There are 21 response variables: `self_employed` to `obs_consequence`, each corresponding to a question on the survey form. We can call `ggplot()` for each variable and examine each plot individually, but we can also iterate this process over multiple variables since we know that they are all `factor`s and we had reordered the response levels during pre-processing.

First, we define the function `plot_by_name()`, it takes two arguments: 

* `col_name` which is the name of the variable to be plotted (a string),  
* `data` which is the dataset to which the variable belongs (a data frame)

The output will be a **bar plot** counting the number of each response in the variable.
 
```{r define function}
plot_by_name <- function(col_name, data){
    ggplot(data, aes_string(col_name)) +
        geom_bar()
}
```

We can use `lapply()` to iterate the above function over (a subset of) the variable names in our dataset, and then use `grid.arrange()` to display all the plots in the list at once.

```{r bar plots 1, fig.width = 12, fig.height = 9}
survey_bar1 <- names(survey_clean)[6:14] %>% 
    lapply(plot_by_name, data = survey_clean) %>% 
    grid.arrange(grobs = ., ncol = 3) # pass the list to the grobs parameter
```

```{r bar plots 2, fig.width = 12, fig.height=9}
survey_bar2 <- names(survey_clean)[15:23] %>% 
    lapply(plot_by_name, data = survey_clean) %>% 
    grid.arrange(grobs = ., ncol = 3)
```

```{r bar plots 3, fig.width = 12, fig.height = 3}
survey_bar3 <- names(survey_clean)[24:26] %>% 
    lapply(plot_by_name, data = survey_clean) %>% 
    grid.arrange(grobs = ., ncol = 3)
```

### Initial observations

* Slightly more than half the respondents indicated they have sought treatment for a mental health condition. (Note that not all respondents may have such a condition)  
* In general, respondents do not indicate high awareness of whether their employers offer benefits, programs or resources for mental health conditions and treatment (`benefits`, `care_options`, `wellness_program`, `seek_help`, `anonymity`, `leave`).  
* Significantly more respondents believed that discussing a mental health issue with their employer would have negative consequences relative to discussing a physical health issue with their employer.  
* However, many respondents appear to be comfortable discussing mental health issues with their coworkers or supervisor (is this implied to be in an informal setting?).
* This stands in contrast with a clear reluctance to bring up mental health issues in an interview with a potential employer.
